---
title: "Week 1: Getting Set Up"
description: "First impressions, and getting the workflow running."
date: 2026-08-28
---

## What We Did

<!-- CHANGE THIS: replace with your own notes -->

Set up the portfolio workflow: made a repository from a template, cloned it with
GitHub Desktop, edited files in RStudio, rendered with Quarto, and published to
GitHub Pages.

## Key Concepts

Three places a file can live:

1. **My computer** — where I edit
2. **GitHub.com** — where it's stored
3. **My live site** — what the world sees

The loop: **Edit → Save → Render → Commit → Push → wait → refresh**

## 0828 Notes
What is a website?
 A folder of html files -> Quarto (code chunck + text)
 + A computer that is always on -> Github Pages
 
 Quarto 
 Qmd (you wirte) -> quarto (machine) -> html (web page)
                    Render
what ahppens in the machine?
1. read files from top to bottom
2. finds code chunks
3. send them to R, get results back
4. glues texts + results into html

anatomy of a .qmd
#--- (no space, 3 dashes)
title: "
date: 2026-08-28
#---
YAML: a mark up langauge  

Markdown: .md

### what i learned today (title)
I learned that the logic of how Qmd and html works for human and machines respectively. I learn the difference between Git, Github, and Github desktop. I know the importance of having a clean and organized folder structure.

code chuncks 
```{r}
library(tidycensus)
```

files + folders:
who is this for? 
 for me (a page/wrote)
 - index.qmd
 - about.qmd
 - weekly_notes.qmd
 - assignment.qmd

For the machine:
 - quarto.yml (setting for whole website
 - freeze
 
 Rproj -> Rstudio
 .gitignore -> git
 Renviron -> R (secret)
 doc -> the link to website
 
 "index" 
 - website rule: as for a folder you get that folders index.html
 your github.io -> docs/index.html
 your github.io/weekly_notes -> docs, weekly nots/index.html
 
 GIT/github.com/github desktop
 
 GIT: version trackers, invisible, runs on your computer
 github.com: a website that stores copy of your folder
 github desktop: buttons that tells git what to do
 
 commit -> "save a labeled snapshots/links to your computer
 push -> save things to github
 
 rscript -> only code
 
 getwd() -> get your directory 


## Proving R Works

The chunk below runs when this page renders. If you can see a result underneath
it on your published site, R and Quarto are talking to each other correctly.

```{r}
sessionInfo()$R.version$version.string
```

## Reflection

<!-- CHANGE THIS: 3-4 sentences -->

What I want out of this course: more coding skills, real-life urban data, and solve meaningful problems.

One thing I'm nervous about: the workload of this class

## Questions I Still Have

- not for now
-
## Notes
---
title: "0828"
output: html_document
date: "2026-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

