---
title: "0828"
output: html_document
date: "2026-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0828 Notes
What is a website?
 A folder of html files -> Quarto (code chunck + text)
 + A computer that is always on -> Github Pages
 
 Quarto 
 Qmd (you wirte) -> quarto (machine) -> html (web page)
                    Render
what ahppens in the machine?
1. read files from top to bottom
2. finds code chunks
3. send them to R, get results back
4. glues texts + results into html

anatomy of a .qmd
--- (no space, 3 dashes)
title: "
date: 2026-08-28
---
YAML: a mark up langauge  

Markdown: .md

### what i learned today (title)
your writing

code chuncks 
```{r}
library(tidycensus)
```

files + folders:
who is this for? 
 for me (a page/wrote)
 - index.qmd
 - about.qmd
 - weekly_notes.qmd
 - assignment.qmd

For the machine:
 - quarto.yml (setting for whole website
 - freeze
 
 Rproj -> Rstudio
 .gitignore -> git
 Renviron -> R (secret)
 doc -> the link to website
 
 "index" 
 - website rule: as for a folder you get that folders index.html
 your github.io -> docs/index.html
 your github.io/weekly_notes -> docs, weekly nots/index.html
 
 GIT/github.com/github desktop
 
 GIT: version trackers, invisible, runs on your computer
 github.com: a website that stores copy of your folder
 github desktop: buttons that tells git what to do
 
 commit -> "save a labeled snapshots/links to your computer
 push -> save things to github
 
 rscript -> only code
 
 getwd() -> get your directory 