---
title: "Week 1: Getting Set Up"
description: "First impressions, and getting the workflow running."
date: 2026-08-28
---

## What We Did

<!-- CHANGE THIS: replace with your own notes -->

Set up the portfolio workflow: made a repository from a template, cloned it with
GitHub Desktop, edited files in RStudio, rendered with Quarto, and published to
GitHub Pages.

## Key Concepts

Three places a file can live:

1. **My computer** — where I edit
2. **GitHub.com** — where it's stored
3. **My live site** — what the world sees

The loop: **Edit → Save → Render → Commit → Push → wait → refresh**

## 0828 Notes
What is a website?
 A folder of html files -> Quarto (code chunck + text)
 + A computer that is always on -> Github Pages
 
 Quarto 
 Qmd (you wirte) -> quarto (machine) -> html (web page)
                    Render
what ahppens in the machine?
1. read files from top to bottom
2. finds code chunks
3. send them to R, get results back
4. glues texts + results into html

anatomy of a .qmd
#--- (no space, 3 dashes)
title: "
date: 2026-08-28
#---
YAML: a mark up langauge  

Markdown: .md

### what i learned today (title)
Today, I learned about the logic behind how QMD and HTML work for humans and machines, respectively. QMD is easier for humans to read and write because it allows us to combine text, code, and formatting in one document. HTML, on the other hand, is the language that browsers use to understand and display the content.

I also learned the differences between Git, GitHub, and GitHub Desktop. Git is a version control system that helps track changes in a project. GitHub is an online platform where repositories can be stored and shared, while GitHub Desktop provides a more visual and user-friendly way to work with Git and GitHub.

Another important thing I learned today is the importance of maintaining a clean and organized folder structure. As projects become more complicated, it can be difficult to find files or understand which files are being used. Keeping folders organized from the beginning can make it easier to manage a project, avoid confusion, and collaborate with others.

Overall, today's lesson helped me better understand how different tools and systems work together behind the scenes to support a digital project.

code chuncks 
```{r}
library(tidycensus)
```

files + folders:
who is this for? 
 for me (a page/wrote)
 - index.qmd
 - about.qmd
 - weekly_notes.qmd
 - assignment.qmd

For the machine:
 - quarto.yml (setting for whole website
 - freeze
 
 Rproj -> Rstudio
 .gitignore -> git
 Renviron -> R (secret)
 doc -> the link to website
 
 "index" 
 - website rule: as for a folder you get that folders index.html
 your github.io -> docs/index.html
 your github.io/weekly_notes -> docs, weekly nots/index.html
 
 GIT/github.com/github desktop
 
 GIT: version trackers, invisible, runs on your computer
 github.com: a website that stores copy of your folder
 github desktop: buttons that tells git what to do
 
 commit -> "save a labeled snapshots/links to your computer
 push -> save things to github
 
 rscript -> only code
 
 getwd() -> get your directory 


## Proving R Works

The chunk below runs when this page renders. If you can see a result underneath
it on your published site, R and Quarto are talking to each other correctly.

```{r}
sessionInfo()$R.version$version.string
```

## Reflection

<!-- CHANGE THIS: 3-4 sentences -->

What I want out of this course: What I want to get out of this course is more coding skills and a better understanding of how coding can be applied to real-world urban data. I hope to become more comfortable using programming tools and learn how to use data to explore, analyze, and visualize urban issues. More importantly, I want to work on meaningful problems rather than just completing technical exercises. I hope this course will help me see how coding and data analysis can be used to better understand cities and potentially contribute to solving real-world planning challenges.

One thing I'm nervous about: One thing I am nervous about is the workload of this class. Learning coding can sometimes be time-consuming, especially when something does not work and it takes a long time to figure out why. Since I am still building my technical skills, I am a little worried about keeping up with the assignments and balancing this class with my other coursework. However, I also hope that by practicing consistently throughout the semester, I will become more comfortable with coding and gain confidence over time.

## Questions I Still Have

- not for now
-
## Notes
---
title: "0828"
output: html_document
date: "2026-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

